---
title: "path2path: directional pathway-by-pathway maps from Perturb-seq"
author: "Jun Li"
package: path2path
bibliography: references.bib
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{path2path: directional pathway maps from Perturb-seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.crop = FALSE)
set.seed(1)
```

# What path2path computes

A genome-scale Perturb-seq screen is, in effect, thousands of parallel
differential-expression experiments: each perturbed gene yields a signed
per-gene statistic across all measured genes, and the results stack into a
perturbation-by-gene matrix $Z$. `path2path` turns this matrix into a
**directional pathway-by-pathway interaction map** $\mathbf{M}$ by running
preranked GSEA twice [@subramanian2005; @korotkevich2019]:

* **Stage 1** scores every *row* of $Z$ (one perturbation) against a pathway
  collection, giving each perturbation a pathway-response profile;
* **Stage 2** ranks the perturbations along each pathway *column* of the
  Stage 1 matrix and scores, for every pathway $j$, whether perturbations of
  $j$'s members concentrate at either end.

Entry $\mathbf{M}_{jk} > 0$ means that perturbations targeting members of
pathway $j$ are enriched toward higher Stage 1 scores for pathway $k$. In a
loss-of-function screen, when those scores reflect decreased expression of
pathway $k$ after perturbation, a positive entry suggests that $j$ supports or
activates the transcriptional program represented by $k$; a negative entry
suggests suppression. The map therefore identifies candidate activating and
suppressive relationships. It is asymmetric because $\mathbf{M}_{jk}$ and
$\mathbf{M}_{kj}$ summarize different perturbation-readout combinations.

Three safeguards are built in: the perturbed gene is excluded from its own
ranked list (removing an on-target tautology that otherwise dominates the
self-interaction diagonal), effective-size filters are applied in both
stages, and entries whose pathway pair overlaps at Jaccard $> 0.5$ are
masked.

# A five-minute tour on synthetic data

The package ships a small simulated dataset with two planted relations:
PATHWAY_A activates PATHWAY_B, and PATHWAY_C suppresses PATHWAY_D.

```{r demo}
library(path2path)
data(demo_z)
data(demo_pathways)
dim(demo_z)
```

Build the map (a serial backend keeps this vignette deterministic; on real
data use the default `BiocParallel::bpparam()`):

```{r pipeline}
res <- path2path(demo_z, demo_pathways,
    stage1 = list(nperm = 200, min_size = 5),
    stage2 = list(min_pu = 5),
    BPPARAM = BiocParallel::SerialParam(RNGseed = 1),
    verbose = FALSE)
round(res$M, 2)
```

Both planted relations are recovered with the right sign and direction, and
the reverse entries are untestable (B and D members were not perturbed, so
those rows fall below the Stage 2 universe filter):

```{r planted}
res$M["PATHWAY_A", "PATHWAY_B"]   # activation, > 0
res$M["PATHWAY_C", "PATHWAY_D"]   # suppression, < 0
res$padj["PATHWAY_A", "PATHWAY_B"]
```

Plot the map with significance dots:

```{r plot, fig.width = 6, fig.height = 5}
plot_map(res, padj_cutoff = 0.05)
```

And inspect *why* one entry is strong, at the level of the underlying
ranking:

```{r entry, fig.width = 6, fig.height = 3.5}
plot_entry(res, demo_pathways, "PATHWAY_A", "PATHWAY_B")
```

# Real data

For a real screen, the input is the perturbation-by-gene matrix of
differential-expression statistics (a DESeq2 Wald $z$, a $z$-normalized
pseudobulk difference, or a signed log-fold change), with rows named by the
perturbed gene symbols:

```{r real, eval = FALSE}
z <- read_input_matrix("zscore_myscreen.csv.gz")
pathways <- load_pathways("hallmark")     # or "reactome", "kegg_legacy", ...
res <- path2path(z, pathways)
write_result(res, dir = "results", tag = "myscreen",
    collection = "hallmark")
```

If the statistics are stored in a `SummarizedExperiment` assay with genes in
rows and perturbations in columns, use `as_path2path_matrix()` to transpose
and validate the assay. `as_path2path_se()` converts a completed map back to a
`SummarizedExperiment` whose map and p-value matrices are assays.

Reproducibility: `BiocParallel` backends draw worker seeds from entropy
unless given one explicitly, so `set.seed()` alone is not enough --- pass
`BPPARAM = MulticoreParam(RNGseed = 1)` (or `SerialParam(RNGseed = 1)`)
to make a run exactly repeatable.

Orientation: statistics are negated internally for loss-of-function screens
(CRISPRi/knockout, the default); set `stage1 = list(negate = FALSE)` for
gain-of-function screens (CRISPRa, overexpression).

# Two routes to significance

`path2path()` attaches **Route A** p-values: the Stage 2 nominal p-values
(`fgsea::fgseaMultilevel`) with one Benjamini-Hochberg family over all
populated entries. On the datasets analyzed in the accompanying manuscript,
the realized false-discovery rate of BH screening at 0.05 remained close to
the nominal level across datasets and pathway collections, and the null
distribution of the nominal p-values was close to uniform over the range that
could be assessed by permutation.

For *individually examined* strong entries, **Route B** reruns the full
pipeline on row/column-permuted inputs, pools the null $|\mathbf{M}'|$ values,
and extrapolates the extreme tail with a generalized-Pareto fit
[@knijnenburg2009]:

```{r routeb, eval = FALSE}
cal <- calibrate_pvalues(res, demo_z, demo_pathways, B = 50, seed_base = 42)
cal$gpd[c("u", "xi", "beta", "n_exc")]
cal$pval_gpd["PATHWAY_A", "PATHWAY_B"]
```

```{r calplot, eval = FALSE, fig.width = 7, fig.height = 3.5}
plot_calibration(cal)
```

`seed_base` makes each input permutation reproducible. For exact
reproducibility of the full null runs, also supply seeded `BPPARAM` objects;
`null_dir` persists completed null maps so interrupted runs can resume. Route A
provides the computationally lighter default for routine screening, while
Route B carries uncertainty from both stages through the permutation.

# Session info

```{r sessioninfo}
sessionInfo()
```

# References
